---
output: github_document
---

<!-- README.md is generated from README.Rmd. Edit THIS file, then knit:
     devtools::build_readme()  (or knitr::knit("README.Rmd"))          -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dpi = 110
)
```

# geoscales

<!-- badges: start -->
[![R-CMD-check](https://github.com/optimal2050/geoscales/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/optimal2050/geoscales/actions/workflows/R-CMD-check.yaml)
[![test-coverage](https://github.com/optimal2050/geoscales/actions/workflows/test-coverage.yaml/badge.svg)](https://github.com/optimal2050/geoscales/actions/workflows/test-coverage.yaml)
[![lint](https://github.com/optimal2050/geoscales/actions/workflows/lint.yaml/badge.svg)](https://github.com/optimal2050/geoscales/actions/workflows/lint.yaml)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

> Nested regions and spatial hierarchies for optimization and simulation models.

`geoscales` is the **spatial-domain** package of the optimal2050 modeling stack
and the spatial companion to [`timescales`](https://github.com/optimal2050/timescales).

This is a **multi-language** project. The R package is the current focus
(Phase 1); a C++ core (Phase 2) and a Python port (Phase 3) are planned.

## Quick demo

A `Geoscale` is a nested spatial partition: a flat table of weighted
leaf regions ("atoms") plus the ordered geoframes that group them.
Aggregation and disaggregation are ONE verb — direction comes from the
geoframe ranks, and totals are preserved either way:

```{r demo-table}
library(geoscales)

gs <- geoscale_from_leaftable(
  data.frame(
    country = c("N", "N", "N", "N", "S", "S"),
    state   = c("N1", "N1", "N2", "N2", "S1", "S1"),
    atom    = paste0("A", 1:6),
    km2     = c(100, 200, 300, 400, 500, 600)
  ),
  geoframes = c("country", "state", "atom"),
  name = "demo"
)

cap <- data.frame(atom = paste0("A", 1:6), capacity = 1:6)
cap |> recast_geoscale(gs, from = "atom", to = "country", rule = "sum")

# ... and back down, split by area
data.frame(country = c("N", "S"), capacity = c(10, 20)) |>
  recast_geoscale(gs, from = "country", to = "state",
                  rule = "sum", weight = "km2")
```

With geometry attached, values land on a map as one ggplot2 layer
(`geom_geoscale()`); without any geometry, `plot()` draws the hierarchy
itself:

```{r demo-map, fig.width = 7, fig.height = 2.8, message = FALSE}
library(ggplot2)
sq <- function(x, y) sf::st_polygon(list(cbind(
  c(x, x + 1, x + 1, x, x), c(y, y, y + 1, y + 1, y))))
gs <- attach_geometry_geoscale(gs, sf::st_sfc(
  sq(0, 1), sq(0, 0), sq(1, 1), sq(1, 0), sq(2, 1), sq(2, 0)))

cap |>
  recast_geoscale(gs, from = "atom", to = "state", rule = "sum") |>
  ggplot() +
  geom_geoscale(gs = gs, z = "capacity", geoframe = "state",
                colour = "white", linewidth = 0.6) +
  scale_fill_viridis_c(option = "G") +
  labs(title = "Capacity recast to states, on the map", fill = NULL) +
  theme_geoscale()
```

```{r demo-icicle, fig.width = 7, fig.height = 2.2, message = FALSE}
plot(gs)
```

See `vignette("geoscales")` for the 5-minute tour, and the
[visualization article](https://optimal2050.github.io/geoscales/r/articles/visualization.html)
for choropleths, recasts seen on the map, and a real-map tour of
Iceland with data attached.

Two things about space that the time domain does not have to deal with, and
which shape the whole design:

* **Geoframes need not nest.** In India's IDEEA region table, the `reg32` code
  `APY` merges Andhra Pradesh with *part of* Puducherry, so `reg35` does not
  nest inside `reg32`. Every conversion therefore routes through the atom
  layer, and cross-cutting geoframes work without special handling.
* **Region codes are not unique across geoframes.** 46 of 62 IDEEA codes appear
  at more than one geoframe (`AN` at seven). So `geoframe` is a required
  argument everywhere — nothing is inferred from a bare code.

### No bundled maps

`geoscales` ships integration code, not data: no boundaries are bundled,
and no map source is selected for you. Providers (Natural Earth via
`ne_geoscale()`, or your own via `register_geo_provider()`) pass a point
of view through and record it in `@meta`.

## Documentation

- **[Project site](https://optimal2050.github.io/geoscales/)** — entry point
  for all language flavours
- **[R reference and articles](https://optimal2050.github.io/geoscales/r/)**

## Status

🚧 In development — pre-1.0, APIs may still change between minor
versions. Feedback and issues are welcome.

## Installation

```r
# From GitHub
remotes::install_github("optimal2050/geoscales")
```

Or via [r-universe](https://optimal2050.r-universe.dev/):

```r
install.packages("geoscales", repos = "https://optimal2050.r-universe.dev")
```

## Repository layout

```
geoscales/
├── DESCRIPTION, NAMESPACE, R/, man/, tests/, vignettes/   # R package (root)
├── inst/include/geoscales/                                # C++ headers (Phase 2)
├── cpp/                                                   # standalone C++ core (Phase 2)
├── python/                                                # Python package (Phase 3)
├── docs/                                                  # unified Quarto site
├── specs/                                                 # cross-language golden tests
└── .github/workflows/                                     # CI
```

## License

Apache-2.0. See [LICENSE](LICENSE).
