---
title: "Data structures"
description: "Anatomy of a Geoscale: the leaftable, geoframes, members, geometry and meta properties, what the validator guarantees, and the rule/map/provider registries."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data structures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(geoscales)
```

## The Geoscale object

A `Geoscale` is an S7 object with five properties:

```{r}
gs <- geoscale_example()
S7::prop_names(gs)
```

| property | type | holds |
|---|---|---|
| `@leaftable` | `data.frame` | one row per **atom**: one column per geoframe, the unique `region` key, weight columns |
| `@geoframes` | `character` | the hierarchy, ordered coarsest first |
| `@members` | named `list` | the ordered code vocabulary at each geoframe |
| `@geometry` | `sfc` or `NULL` | optional atom polygons, aligned to leaftable rows |
| `@meta` | `list` | `name`, `desc`, `weights`, `default_weight`, `crs`, `source`, `labels` |

Everything else — parent/child tables, ancestry, shares, dissolved
shapes — is derived on demand; nothing is cached on the object.

## The leaftable

```{r}
geoscale_leaftable(gs)
```

Each row is one atom. The `region` column is the **authored node ID**
(unique, non-missing); geoframe columns may hold `NA` — an atom with no
code at a coarser geoframe (a rest-of-world row, offshore areas) simply
does not participate at that resolution, and `na_action=` decides what
recasts do about it.

## Geoframes, members, ranks

```{r}
geoscale_geoframes(gs)
geoscale_geoframes(gs, finest = TRUE)   # the atom geoframe
geoscale_regions(gs, "state")           # the members at one geoframe
geoscale_rank(gs, c("zone", "atom"))    # positions, coarsest = 1
```

Geoframe names must be syntactically valid column names;
`CORE_GEOFRAMES` is recommended vocabulary, not enforcement:

```{r}
CORE_GEOFRAMES
is_valid_geoframe(c("COUNTRY", "reg32", "", "2bad"))
```

Navigation is derived on demand from the leaftable — and because
geoframes may cross-cut, ancestry is computed atom-mediated, never as a
transitive closure:

```{r}
geoscale_parents(gs, "zone", "ZB", to = "state")   # cross-cut: two parents
geoscale_ancestors(gs, "atom", "A5")
geoscale_descendants(gs, "country", "N")
head(geoscale_ancestry(gs), 4)
```

## Weights and meta

```{r}
geoscale_weights(gs)
gs@meta[c("name", "weights", "default_weight")]
```

`meta$name` is load-bearing: conversion and attach name their working
columns after the object (`geoscale_map()` label columns,
`join_geoscale()`'s `"<name>."` prefixes), so both require a named
Geoscale. `geoscale_build()` derives a default name from the geoframes;
`geoscale_from_leaftable()` takes `name=`.

## What the validator guarantees

Construction (any layer: `geoscale_from_provider()`,
`geoscale_build()`, `geoscale_from_leaftable()`) enforces:

- `region` present, character, unique, no missing/empty values;
- every geoframe present as a leaftable column; members exactly the
  non-`NA` values of that column, in a stable order;
- weights numeric, finite, non-negative, with a positive sum;
  `default_weight` one of the declared weights;
- geometry, when attached, aligned 1:1 with leaftable rows (see
  `attach_geometry_geoscale()`; the CRS is recorded in `meta$crs`);
- a warning (not an error) when geoframes look ordered finer-to-coarser
  — order carries meaning: aggregation direction follows it.

The old argument names `leaves=`/`levels=` are rejected with a pointer
to the current spelling.

## The registries

Three package-level registries change behavior explicitly — the only
way anything global changes:

**Rules** map value-column names to their aggregation rule (there is no
fallback — unregistered columns without `rule=` error):

```{r}
GEOSCALE_RULES
register_geoscale_rule("capacity", "sum")
register_geoscale_rule("eff", "weighted_mean", weight = "pop")
get_geoscale_rule("eff")
list_geoscale_rules()
clear_geoscale_rules()
```

**Maps** hold exact crosswalks that short-circuit the atom-layer
derivation in `geoscale_map()` — for hand-audited concordances:

```{r}
fake <- data.frame(state = "N1", zone = "ZC", n_from = 1L,
                   n_overlap = 1L, w = 1, w_from = 1)
register_geoscale_map("state", "zone", fake, gs = gs)
list_geoscale_maps()
get_geoscale_map("state", "zone", gs = gs)
clear_geoscale_maps()
```

**Providers** are named map sources (`fetch` functions returning a wide
`sf`/`data.frame`); the package ships one:

```{r}
list_geoscale_providers()
get_geoscale_provider("naturalearth")$desc
```

## Where to next?

- [Concepts](concepts.html) — why the structures look this way.
- [Data manipulation](data-manipulation.html) — the verbs over these
  structures.
- [Building from Natural Earth](from-naturalearth.html) — providers in
  practice.
