---
title: "Data manipulation with geoscales"
description: "The tidy workflow end to end: attach a Geoscale to region-keyed tables, recast values between resolutions, inspect the crosswalk, chain time and space with recast(), and run the same pipelines over data.table and arrow backends."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data manipulation with geoscales}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(geoscales)
library(dplyr, warn.conflicts = FALSE)
has_dt <- requireNamespace("data.table", quietly = TRUE) &&
  requireNamespace("dtplyr", quietly = TRUE)
```

## The toolkit at a glance

| verb | direction | what it does |
|---|---|---|
| `join_geoscale()` | Geoscale → columns | *attach* labels, membership, share/weight to a table (no aggregation) |
| `recast_geoscale()` | geoframe A → geoframe B | *convert* values between resolutions, one rule per column |
| `recast_to_geoatoms()` / `recast_from_geoatoms()` | the route halves | project down to the atom layer / aggregate up from it |
| `geoscale_map()` | A → B crosswalk | the conversion, materialised as a small table |
| `register_geo_rule()` / `register_geo_map()` / `register_geo_provider()` | registries | per-column rules, exact crosswalks, map sources |
| `recast()` | generic | one pipeline verb across time AND space |

Everything below runs on the synthetic `geoscale_example()` fixture in
tidyverse style — data flows through `|>`, and every verb accepts a
`data.frame`, tibble, `data.table`, dtplyr, or arrow input (see
[Backends](#backends)).

```{r}
gs <- geoscale_example()
atoms <- gs@leaftable$region[!is.na(gs@leaftable$country)]
cap <- tibble(atom = atoms, capacity = c(1, 2, 3, 4, 5, 6))
```

## Attaching a Geoscale to a table

`join_geoscale()` *decorates* rather than converts: it adds a label
column **named after the Geoscale**, plus optional coarser-geoframe
membership columns and share/weight, all `"<name>."`-prefixed. The keyed
geoframe is inferred from the columns (or passed as `geoframe=`):

```{r}
tibble(state = c("N1", "N2", "S1"), v = 1:3) |>
  join_geoscale(gs, geoframes = TRUE, meta = TRUE)
```

Because every Geoscale attaches under its own name, several can coexist
on one dataset — and a table carrying two label columns is itself an
empirical crosswalk between the region systems. Existing columns are
never overwritten; a clashing attach errors instead.

## Recasting between geoframes

`recast_geoscale()` converts values — aggregation and disaggregation are
one operation, routed through the atom layer, so cross-cutting
geoframes need no special handling. Three things to know:

**1. One rule per value column, and the rule is mandatory.** Pass
`rule=` for all columns, or register per-column rules once; a column
with neither errors — a silently guessed rule is a silent unit error.

```{r}
register_geo_rule("capacity", "sum")            # extensive
register_geo_rule("eff", "weighted_mean",       # intensive
                  weight = "pop")

cap |> recast_geoscale(gs, from = "atom", to = "country")
```

| rule | up (coarsen) | down (refine) |
|---|---|---|
| `sum` | sum | split by the weight (conserves totals) |
| `weighted_mean` | weight-weighted mean | copy |
| `mean` | plain mean | copy |
| `copy` | common value (error if not constant) | copy |
| `sd` | dispersion over atoms | degenerates |

Disaggregation is the same call in the other direction — the chosen
weight drives the split, and a round trip is exact:

```{r}
tibble(country = c("N", "S"), capacity = c(10, 20)) |>
  recast_geoscale(gs, from = "country", to = "state", weight = "km2")
```

**2. Identifier columns ride along.** Columns that are neither the key
nor values (a year, a technology) are grouping columns, so panel data
converts in one call, and totals conserve *per group*. Materialised
results complete to the **full target vocabulary** in member order
(`NA` where nothing landed), so downstream joins see a stable schema.

```{r}
panel <- expand.grid(atom = atoms, year = c(2030L, 2050L),
                     stringsAsFactors = FALSE) |>
  as_tibble() |>
  mutate(capacity = seq_along(atom) * 1.0)
panel |> recast_geoscale(gs, from = "atom", to = "country",
                         values = "capacity")
```

(`values=` is explicit here because `year` is numeric — value
auto-detection takes every numeric non-key column, so numeric
identifiers must be named out.)

**3. Coverage is explicit.** Atoms with no code at `from` or `to` go
through `na_action=`: `"drop"` (default; warns — the affected share is
genuinely lost), `"keep"` (an explicit `NA` region row so totals
conserve), or `"error"`. Note that downstream, energyRt reads `NA` in a
region column as a wildcard meaning *all* regions — don't pass `"keep"`
output there unfiltered.

```{r, include = FALSE}
clear_geo_rules()
```

## The route halves and cross-Geoscale conversion

`recast_geoscale()` is the fused route `from -> atoms -> to`; its halves
are public. Going down, extensive columns split by the chosen weight and
the atom `weight` is attached so the return trip is exact; going up,
rules act on the atom rows directly:

```{r}
a <- tibble(country = c("N", "S"), capacity = c(10, 20)) |>
  recast_to_geoatoms(gs, from = "country", rule = "sum", weight = "km2")
head(a, 3)
a |> recast_from_geoatoms(gs, to = "state", rule = "sum")
```

Because the atom rows are keyed by atom `region` IDs, the halves also
convert **across two different Geoscales that share atom keys** (reg32
to NUTS style conversions) — and the fused verb accepts another
Geoscale as `to`, targeting its atom layer:

```{r}
lf <- tibble(band = rep(c("X", "Y"), 3), atom = atoms,
             km2 = c(100, 200, 300, 400, 500, 600))
gs_b <- geoscale_from_leaftable(lf, geoframes = c("band", "atom"),
                                name = "bands")
cap |>
  recast_geoscale(gs, from = "atom", to = gs_b, rule = "sum") |>
  recast_geoscale(gs_b, from = "atom", to = "band", rule = "sum")
```

## The crosswalk, inspectable and overridable

Every recast is a join against a small crosswalk table — one row per
overlapping region pair with atom counts (`n_from`, `n_overlap`) and
weights (`w`, the absolute overlap weight; `w_from`, the full source
weight — `w / w_from` is the split share):

```{r}
geoscale_map("country", "state", gs = gs, weight = "km2")
```

When an exact correspondence is known (an official concordance table),
register it and it short-circuits the derivation — with accessors to
inspect what is installed:

```{r}
exact <- geoscale_map("state", "zone", gs = gs)
register_geo_map("state", "zone", exact, gs = gs)
list_geo_maps()
identical(get_geo_map("state", "zone", gs = gs), exact)
clear_geo_maps()
```

## One verb across time and space

The bare `recast()` generic (owned by timescales, re-exported here)
dispatches on the scale object, so one pipeline chains both dimensions —
this runs, since geoscales imports timescales:

```{r}
library(timescales)
x <- expand.grid(timeslice = sprintf("m%02d", 1:12),
                 state = c("N1", "N2", "S1"),
                 stringsAsFactors = FALSE) |>
  as_tibble() |>
  mutate(demand = seq_along(timeslice) * 1.0)

x |>
  recast(calendar("m12"), to = calendar("q4"), year = 2025,
         rule = "sum") |>
  recast(gs, to = "country", rule = "sum")
```

Same contract on both sides: explicit rules, identifier columns
preserved, totals conserved under `"sum"`.

## Backends

The verbs are single dplyr pipelines, so the SAME code runs over an
in-memory `data.frame`/tibble, a `data.table` (via dtplyr), or an arrow
Dataset/query. Eager inputs come back in their own class; lazy inputs
return the *uncollected query* unless `collect = TRUE`:

```{r, eval = has_dt}
dt <- data.table::as.data.table(cap)

dt |>
  recast_geoscale(gs, from = "atom", to = "country", rule = "sum") |>
  class()                          # data.table in, data.table out

lazy <- dtplyr::lazy_dt(dt) |>
  recast_geoscale(gs, from = "atom", to = "country", rule = "sum")
class(lazy)                        # the query, not the result
as.data.frame(dplyr::collect(lazy))
```

Two contract details for lazy sources (dtplyr, arrow): results carry the
*observed* target regions only — the full-vocabulary completion happens
on materialisation — and the geoscale side of every join is a small
in-memory frame, so an on-disk arrow dataset is never pulled into memory
for the geoscale arithmetic.

## Where to next?

- [Concepts](concepts.html) — why the route always goes through the
  atom layer.
- [Data structures](data-structures.html) — the registries as
  structures.
- [Visualization](visualization.html) — the same pipelines flowing onto
  maps.
