---
title: "Building a Geoscale from Natural Earth"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building a Geoscale from Natural Earth}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
has_ne <- requireNamespace("rnaturalearth", quietly = TRUE) &&
  requireNamespace("rnaturalearthdata", quietly = TRUE) &&
  requireNamespace("sf", quietly = TRUE)
has_hires <- has_ne && requireNamespace("rnaturalearthhires", quietly = TRUE)
knitr::opts_chunk$set(eval = has_ne)
```

```{r setup, eval = TRUE}
library(geoscales)
```

## The package ships no maps

`geoscales` contains integration code, not data. There is no `data/`
directory and no bundled boundaries. A provider fetches a source table; the
package records what was fetched.

```{r, eval = TRUE}
geo_list_providers()
```

Natural Earth is the default provider because its country table is *already*
a wide leaves table — one `sf` object carries the whole nest as columns:

```
continent (8) -> region_un -> subregion (22) -> sovereignt
              -> admin / adm0_a3 (177) -> geounit -> subunit
```

with `pop_est` and `gdp_md` as ready-made weights.

```{r}
gs <- ne_geoscale(scale = 110)
gs
```

## Aggregating up

We will follow Iceland through the hierarchy. It sits in Northern Europe:

```{r}
lf <- S7::prop(gs, "leaves")
lf[lf$country == "ISL", c("continent", "subregion", "country", "pop_est")]
```

Rolling population up from countries to sub-regions is one call. `rule = "sum"`
because population is *extensive* — it adds up:

```{r}
pop <- data.frame(country = lf$country, pop = lf$pop_est)
pop <- pop[!is.na(pop$country), ]

agg <- geo_recast(pop, gs, from = "country", to = "subregion", rule = "sum")
agg[agg$subregion == "Northern Europe", ]
```

## ... and back down

The same verb runs the other way. Going down, `rule = "sum"` splits a total in
proportion to a weight, so the total is conserved:

```{r}
ne <- agg[agg$subregion == "Northern Europe", ]

back <- geo_recast(ne, gs, from = "subregion", to = "country",
                   rule = "sum", weight = "pop_est")
head(back[order(-back$pop), ], 4)
```

Because we split by the same quantity we aggregated, the round trip is exact:

```{r}
back$pop[back$country == "ISL"]
sum(back$pop) == sum(ne$pop)
```

That is the whole idea: **aggregation and disaggregation are one operation.**
Direction follows the level ranks, and `geo_recast()` routes through the atom
layer either way.

## Iceland's own regions

Natural Earth's admin-1 layer gives sub-country units. This needs
`rnaturalearthhires`, which is **not on CRAN**:

```{r, eval = FALSE}
install.packages("rnaturalearthhires",
                 repos = "https://ropensci.r-universe.dev")
```

```{r, eval = has_hires}
s <- ne_source(level = "states", country = "Iceland")
d <- as.data.frame(s)
nrow(d)
```

Nine units — but Iceland has eight regions. `Reykjavik` and the surrounding
capital area are separate admin-1 features that belong to the same region,
which Natural Earth records in `gn_name`. That gives a genuinely **ragged**
hierarchy: one group with two children, seven with one.

```{r, eval = has_hires}
isl <- data.frame(
  country    = "ISL",
  landshluti = d$gn_name,       # Iceland's eight regions
  unit       = d$iso_3166_2,    # the nine Natural Earth units
  stringsAsFactors = FALSE
)

g <- geoscale_from_leaves(isl, levels = c("country", "landshluti", "unit"),
                          key = "unit", name = "iceland")
g <- geo_attach_geometry(g, s, by = "iso_3166_2", level = "unit")
g <- geo_area(g, name = "km2")
g
```

No level is called `region` here: `geoscale_from_leaves()` uses that name for
the internal atom key, so it is reserved for the finest level.

The Capital Region is the one that has two units under it:

```{r, eval = has_hires}
geo_children(g, "landshluti", "Hofudborgarsvadi")
geo_nests(g, "landshluti", "unit")
```

### Up: units to regions to country

```{r, eval = has_hires}
set.seed(1)
x <- data.frame(unit = geo_regions(g, "unit"),
                generation = round(runif(9, 50, 500)))

geo_recast(x, g, from = "unit", to = "landshluti", rule = "sum")
```

```{r, eval = has_hires}
geo_recast(x, g, from = "unit", to = "country", rule = "sum")
sum(x$generation)
```

### Down: a national total split by area

```{r, eval = has_hires}
tot <- data.frame(country = "ISL", demand = 1000)

dd <- geo_recast(tot, g, from = "country", to = "unit",
                 rule = "sum", weight = "km2")
dd[order(-dd$demand), ]
```

Look at what that produces. The Capital Region units receive almost nothing,
because they are tiny — yet they hold most of Iceland's population and would
consume most of its electricity. **Area is rarely the right weight for
demand.** This is precisely why `geo_recast()` makes you name a `weight =`
rather than picking one for you; with a population column on the leaves you
would pass that instead.

An *intensive* quantity must not be summed at all — use `weighted_mean`, which
copies going down and weight-averages going up:

```{r, eval = has_hires}
cf <- data.frame(unit = geo_regions(g, "unit"),
                 cf = round(runif(9, 0.2, 0.5), 3))

geo_recast(cf, g, "unit", "landshluti", rule = "weighted_mean", weight = "km2")
```

### Geometry follows the levels

`geo_geometry()` dissolves the atoms up to whatever level you ask for, so the
nine units become eight regions:

```{r, eval = has_hires}
shp <- geo_geometry(g, level = "landshluti")
nrow(shp)
```

```{r, eval = has_hires && requireNamespace("ggplot2", quietly = TRUE), fig.width = 7, fig.height = 4}
geo_plot(g, level = "landshluti")
```

## Other providers

The same interface accepts any source. Register one with
`geo_register_provider()`, or pass an `sf` object straight to
`geoscale_from_provider()`. Natural candidates are `giscoR` (authoritative
NUTS for Europe), `geodata`/GADM (admin-2 and admin-3 — note its licence
forbids commercial use) and `tigris` (US Census).
