---
title: "Getting started with geoscales"
description: "A short, end-to-end tour: build a Geoscale, recast values between resolutions, attach it to data, and draw a map."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with geoscales}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(geoscales)
library(dplyr, warn.conflicts = FALSE)
library(ggplot2, warn.conflicts = FALSE)
# pre-built by data-raw/iceland_wind.R; absent from the built package, so
# the wind-cluster section below renders its outputs only on the dev machine
has_wind <- file.exists("../data-raw/iceland_wind.rds") &&
  requireNamespace("ggplot2", quietly = TRUE) &&
  requireNamespace("energypal", quietly = TRUE)
```

## What problem does this solve?

Energy-system, climate, and policy models carve space into discrete
*regions*, and different models pick different carvings: one nation,
five grid regions, thirty-two model regions, forty-six zones. The codes
are arbitrary, the weights (area, population) are data, and region
systems drawn by different hands rarely nest — converting values
between them is where unit errors live.

`geoscales` represents any such carving as a **Geoscale**: a set of
*atoms* (the finest regions) plus ordered *geoframes* that group them —
the spatial companion to a `timescales::Calendar`. With one object you
get:

- a stable schema for region codes and their weights,
- well-defined conversions between any two resolutions — including
  cross-cutting ones, because every conversion routes through the atom
  layer,
- attachment of hierarchy columns to your tables, and ggplot2-ready
  maps.

## A 5-minute tour

### 1. Build a Geoscale

Three construction layers, from most to least convenient — a provider,
parent-child crosswalks, or a wide table you already have:

```{r}
gs <- geoscale_from_leaftable(
  data.frame(
    country = c("N", "N", "N", "N", "S", "S"),
    state   = c("N1", "N1", "N2", "N2", "S1", "S1"),
    atom    = c("A1", "A2", "A3", "A4", "A5", "A6"),
    km2     = c(100, 200, 300, 400, 500, 600)
  ),
  geoframes = c("country", "state", "atom"),
  name = "tour"
)
gs
```

(`geoscale_build()` assembles the same thing from ragged parent-child
crosswalks; `geoscale_from_provider()` pulls a source like Natural
Earth — see `vignette("from-naturalearth")`.)

### 2. Inspect the structure

```{r}
gs@geoframes                 # the hierarchy, coarsest first
head(geoscale_leaftable(gs), 3)        # one row per atom
geoscale_regions(gs, "state")
geoscale_share(gs, "state", weight = "km2", within = "country")
```

### 3. Convert data between resolutions

One rule per value column; aggregation and disaggregation are the same
operation, and totals conserve under `"sum"`:

```{r}
cap <- tibble(atom = paste0("A", 1:6), capacity = c(1, 2, 3, 4, 5, 6))
cap |> recast_geoscale(gs, from = "atom", to = "country", rule = "sum")

# ... and back down, split by area
tibble(country = c("N", "S"), capacity = c(10, 20)) |>
  recast_geoscale(gs, from = "country", to = "state",
                  rule = "sum", weight = "km2")
```

The `rule` is deliberately mandatory — pass one, or register it per
column with `register_geoscale_rule()`; a silently guessed rule would be a
silent unit error.

### 4. Attach a Geoscale to a table

`join_geoscale()` decorates rather than converts — membership and
share/weight columns arrive `"<name>."`-prefixed, so several Geoscales
can coexist on one dataset:

```{r}
tibble(state = c("N1", "N2", "S1"), v = 1:3) |>
  join_geoscale(gs, geoframes = TRUE, meta = TRUE)
```

### 5. Visualize

`geoscale_autoplot()` (also `plot()`) draws the hierarchy itself — no
geometry needed (and with `data =`/`z =` its bands fill with a value
recast to every geoframe — see the visualization article):

```{r, fig.width = 7, fig.height = 3, eval = requireNamespace("ggplot2", quietly = TRUE)}
geoscale_autoplot(gs)
```

With geometry attached (any `sfc` aligned with the atoms),
`geom_geoscale()` puts values on a map inside a normal `ggplot()`
pipeline. The data must be keyed at the geoframe you draw, so recast
finer data up first:

```{r, fig.width = 7, fig.height = 2.8, eval = requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("sf", quietly = TRUE)}
sq <- function(x, y) sf::st_polygon(list(cbind(
  c(x, x + 1, x + 1, x, x), c(y, y, y + 1, y + 1, y))))
gs <- attach_geometry_geoscale(gs, sf::st_sfc(
  sq(0, 1), sq(0, 0), sq(1, 1), sq(1, 0), sq(2, 1), sq(2, 0)))

cap |>
  recast_geoscale(gs, from = "atom", to = "state", rule = "sum") |>
  ggplot() +
  geom_geoscale(gs = gs, z = "capacity", geoframe = "state") +
  scale_fill_viridis_c() +
  theme_geoscale()
```

(The visualization article does the same on real maps.)

## From raster data to a Geoscale: wind clusters

The stack on the package's front page is a real, data-built hierarchy:
Iceland's onshore wind resource from the
[Global Wind Atlas](https://globalwindatlas.info/), clustered *within*
each administrative region into contiguous wind-speed classes. That is
the standard renewable-modeling move — a model wants a handful of
supply regions per admin unit, ranked by resource quality, not two
million raster cells.

The pipeline below is complete and pasteable (packages: sf, terra,
dplyr, [globalwindatlas](https://github.com/energyRt/globalwindatlas);
a one-time ~64 MB download plus a few minutes of geometry work). It is
the same code as `data-raw/iceland_wind.R` in the repository, which
additionally caches the slow steps and pre-saves the result. Not run
while building this page.

**Raw materials** — the wind-speed raster and the admin regions:

```{r, eval = FALSE}
library(sf)     # load sf BEFORE terra -- the reverse order can crash
library(terra)  # an R session on Windows (GDAL DLL clash)
library(dplyr)

# 100 m mean wind speed for Iceland, one GeoTIFF. Pass `filename`
# explicitly -- the default filename handling mangles "data-raw"
tif <- globalwindatlas::gwa_get_wind_speed(
  "ISL", height = 100, filename = "data-raw/gwa/ISL_wind_speed_100.tif")

# onshore regions from Natural Earth; the capital region arrives split
# in two -- merge by name, and keep an ISO code per region for the
# cluster ids
ne  <- ne_source(geoframe = "states", country = "Iceland")
reg <- ne |>
  group_by(region = gn_name) |>
  summarise() |>
  st_make_valid()
iso <- ne |>
  st_drop_geometry() |>
  group_by(region = gn_name) |>
  summarise(iso = max(iso_3166_2))

# contiguous ">= 8" and ">= 10" m/s resource classes within each
# region: threshold -> polygonize -> simplify -> buffer -> drop crumbs
int <- c(0, 8, 10)
grp <- globalwindatlas::gwa_group_locations(
  tif, gis_sf = reg, ID = "region", int = int,
  aggregate_tif = 2, plot_process = FALSE)
```

**Classes to clusters** — everything in a projected CRS (ISN93
Lambert), where the geometry ops are robust. Two boundary rules keep
the later dissolves exact (up-aggregation on the map is a *union* of
atoms, and unions only merge cleanly when shared edges match to the
last coordinate): region borders are simplified *topology-aware*
(`rmapshaper::ms_simplify()` simplifies each shared arc once —
`sf::st_simplify()` would treat each side independently and leave
sliver "ghost borders"), and each region is carved so that every piece
is differenced from the region itself, the last class being the
region's remainder:

```{r, eval = FALSE}
grp_m <- st_as_sf(grp)[, c("region", "int")] |>
  st_transform(3057) |> st_make_valid() |>
  st_simplify(dTolerance = 300) |> st_make_valid()
reg_m <- reg |>
  st_transform(3057) |> st_make_valid() |>
  rmapshaper::ms_simplify(keep = 0.2, keep_shapes = TRUE) |>
  st_make_valid()

only_poly <- function(g) {          # keep polygons, drop stray pieces
  if (length(g) == 0) return(st_sfc(st_polygon(), crs = st_crs(g)))
  g <- st_make_valid(g)
  i <- st_geometry_type(g) == "GEOMETRYCOLLECTION"
  if (any(i)) g[i] <- st_collection_extract(g[i], "POLYGON") |> st_union()
  g
}

clusters <- lapply(seq_len(nrow(reg_m)), function(i) {
  rg  <- st_geometry(reg_m)[i]
  cls <- grp_m[grp_m$region == reg_m$region[i] & grp_m$int > min(int), ]
  cls <- cls[order(-cls$int), ]
  out <- vector("list", nrow(cls) + 1L)
  acc <- NULL                       # union of the pieces cut so far
  for (k in seq_len(nrow(cls))) {
    pk <- only_poly(st_intersection(st_geometry(cls)[k], rg))
    if (!is.null(acc)) pk <- only_poly(st_difference(pk, acc))
    acc <- if (is.null(acc)) pk else only_poly(st_union(acc, pk))
    out[[k]] <- st_sf(region = reg_m$region[i], int = cls$int[k],
                      geometry = pk)
  }
  out[[nrow(cls) + 1L]] <- st_sf(region = reg_m$region[i], int = min(int),
                                 geometry = only_poly(st_difference(rg, acc)))
  do.call(rbind, out)
})

clusters <- do.call(rbind, clusters) |>
  filter(!st_is_empty(geometry),
         as.numeric(units::set_units(st_area(geometry), "km^2")) > 1) |>
  group_by(region) |>
  arrange(desc(int), .by_group = TRUE) |>
  mutate(rank = row_number()) |>   # c1 = windiest class of its region
  ungroup() |>
  left_join(iso, by = "region") |>
  mutate(cluster = paste0(iso, "_c", rank)) |>
  st_as_sf()

# per-cluster mean wind speed (from the full-resolution raster) + area
r <- rast(tif)
clusters$wind <- terra::extract(
  r, vect(st_transform(clusters, st_crs(r))),
  fun = mean, na.rm = TRUE, ID = FALSE)[[1]]
clusters$km2 <- as.numeric(units::set_units(st_area(clusters), "km^2"))
```

**Assemble the Geoscale** — the clusters are the atoms:

```{r, eval = FALSE}
gs <- geoscale_from_leaftable(
  clusters |>
    st_drop_geometry() |>
    transmute(country = "Iceland", landshluti = region, cluster, km2),
  geoframes = c("country", "landshluti", "cluster"),
  key = "cluster", name = "iceland_wind"
) |>
  attach_geometry_geoscale(clusters, by = "cluster", geoframe = "cluster")

wind <- clusters |> st_drop_geometry() |> select(cluster, wind)
```

The result is an ordinary three-geoframe Geoscale — the clusters are
its atoms, so everything above applies. The repository ships it
pre-built (`data-raw/iceland_wind.rds`, written by
`data-raw/iceland_wind.R`), which is what renders here: recast the
per-cluster wind speed up to the regions with an area-weighted mean,

```{r, eval = has_wind}
iceland <- readRDS("../data-raw/iceland_wind.rds")
head(geoscale_leaftable(iceland$gs), 4)

iceland$wind |>
  recast_geoscale(iceland$gs, from = "cluster", to = "landshluti",
                  rule = "weighted_mean", weight = "km2")
```

or hand the atom-level values straight to the stack view, which recasts
them onto every plane itself (`data`/`z`) and labels a chosen geoframe
(`labels`). `palette = NULL` leaves the fill scale to the caller —
here the Global Wind Atlas palette from
[energypal](https://github.com/optimal2050/energypal), whose colours
sit on the atlas's absolute breaks (`limits` only windows the legend);
`frame` and `connectors` draw the plane sheets and corner guides that
make the perspective legible around curved coastlines, a mostly
transparent `frame_fill` turns the sheets into glass panes, and a
per-plane `colour` vector drops the borders on the fragment-heavy
cluster plane. This is the front-page figure:

```{r wind-stack, eval = has_wind, fig.width = 8, fig.height = 3.6, dpi = 130}
geoscale_autoplot(iceland$gs, type = "stack", view = "perspective",
                  direction = "down", data = iceland$wind, z = "wind",
                  labels = "landshluti", palette = NULL, gap = .275,
                  colour = c("grey35", "grey35", NA),  # no cluster borders
                  frame = TRUE, connectors = FALSE,
                  frame_fill = ggplot2::alpha("#6FA8DC", 0.15)) +
  energypal::scale_fill_energy_b(limits = c(6.5, 11.5)) +
  ggplot2::labs(fill = "m/s at 100m")
```

*Wind data: [Global Wind Atlas](https://globalwindatlas.info) — DTU,
in partnership with the World Bank Group, data by Vortex, funded by
ESMAP.*

Offshore wind areas would slot in as a sibling branch of the same
hierarchy, but they first need a regionalization of the sea area —
deferred for now.

## Where to next?

- [Concepts](concepts.html) — atoms, partitions vs trees, the shared
  *scales glossary.
- [Data structures](data-structures.html) — anatomy of a `Geoscale`
  and its registries.
- [Data manipulation](data-manipulation.html) — attach, recast, route
  halves, crosswalks, backends, and chaining time with space.
- [Building from Natural Earth](from-naturalearth.html) — providers in
  practice.
- [Visualization](visualization.html) — maps with `geom_geoscale()`,
  on a real multi-layer example.
