---
title: "Getting started with geoscales"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with geoscales}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(geoscales)
```

## What a Geoscale is

A `Geoscale` is a nested spatial partition — the spatial companion to a
`timescales::Calendar`. It holds:

* **`leaves`** — one row per *atom* (the finest region), one column per level,
  a unique `region` key, and one or more named weight columns;
* **`levels`** — the hierarchy, ordered coarsest first;
* **`members`** — the code vocabulary at each level;
* **`meta`** — name, weights, source, CRS.

Everything else — parent/child tables, ancestry, shares — is derived on
demand. Nothing is cached on the object.

```{r}
gs <- geoscale_example()
gs
```

## Levels are partitions, not necessarily a tree

This is the main thing that differs from the time domain. Time levels nest
cleanly: every hour belongs to exactly one day. Spatial levels frequently do
not.

In the example, zone `ZB` draws atoms from two different states, so `state`
and `zone` cross-cut:

```{r}
geo_nests(gs, "country", "state")
geo_nests(gs, "state", "zone")
```

Real data behaves the same way. In the IDEEA region table for India, the
`reg32` code `APY` merges Andhra Pradesh with *part of* Puducherry, so
`reg35` does not nest inside `reg32`.

This is exactly why the atom layer exists, and why `geo_recast()` always
routes through it.

## Codes are not unique across levels

The code `"N1"` exists at both `state` and `zone`. A bare code is therefore
ambiguous, and every lookup must name its level:

```{r}
geo_children(gs, "state", "N1")
geo_children(gs, "zone", "N1")
```

In IDEEA, 46 of 62 codes appear at more than one level — `AN` appears at
seven. This is the norm, not an edge case.

## Recasting: one verb, both directions

`geo_recast()` projects source values down to atoms and then aggregates them
up to the target, so aggregation and disaggregation are the same operation.
The direction falls out of the level ranks.

Going up, an extensive quantity is summed:

```{r}
cap <- data.frame(atom = c("A1", "A2", "A3", "A4", "A5", "A6"),
                  capacity = c(1, 2, 3, 4, 5, 6))
geo_recast(cap, gs, from = "atom", to = "country", rule = "sum")
```

Going down, the same rule splits proportionally to a weight:

```{r}
y <- data.frame(country = c("N", "S"), capacity = c(10, 20))
geo_recast(y, gs, from = "country", to = "state",
           rule = "sum", weight = "km2")
```

Totals are preserved, and a round trip is exact:

```{r}
up   <- geo_recast(cap, gs, "atom", "state", rule = "sum")
back <- geo_recast(up, gs, "state", "atom", rule = "sum", weight = "km2")
back
```

### Rules

| rule | up (coarsen) | down (refine) | use for |
|---|---|---|---|
| `sum` | sum | split by weight | capacity, demand, area |
| `weighted_mean` | weighted mean | copy | efficiency, price, capacity factor |
| `mean` | plain mean | copy | diagnostics |
| `copy` | common value | copy | region-invariant scalars |

An intensive quantity must not be summed:

```{r}
eff <- data.frame(atom = c("A1", "A2", "A3", "A4", "A5", "A6"),
                  eff = c(0.3, 0.4, 0.5, 0.5, 0.6, 0.6))
geo_recast(eff, gs, "atom", "state", rule = "weighted_mean", weight = "pop")
```

### Registering rules per parameter

Rather than passing `rule=` at every call site, register it once. Each value
column is then converted by its own rule in a single call:

```{r}
geo_register_rule("capacity", "sum")
geo_register_rule("eff", "weighted_mean", weight = "pop")

mix <- data.frame(atom = c("A1", "A2", "A3", "A4", "A5", "A6"),
                  capacity = c(1, 2, 3, 4, 5, 6),
                  eff = c(0.3, 0.4, 0.5, 0.5, 0.6, 0.6))
geo_recast(mix, gs, "atom", "state")
```

```{r, include = FALSE}
geo_clear_rules()
```

An explicit `rule=` always overrides the registry.

## Partial coverage

Region tables routinely have atoms with no code at a coarser level — a
rest-of-world row, offshore zones, unassigned territory. The example has an
atom `ROW` with no country.

`na_action` decides what happens to it:

```{r}
x <- rbind(cap, data.frame(atom = "ROW", capacity = 100))

# "drop" (default) loses the uncovered share, and says so
out <- suppressWarnings(
  geo_recast(x, gs, "atom", "country", rule = "sum"))
sum(out$capacity)

# "keep" conserves the total in an explicit NA group
kept <- geo_recast(x, gs, "atom", "country", rule = "sum",
                   na_action = "keep")
sum(kept$capacity)
```

Note that `energyRt` reads `NA` in a region column as *a wildcard meaning all
regions*, so `"keep"` output should not be passed there unfiltered.

## Identifier columns

Columns that are neither the key nor a value are kept as grouping columns, so
panel data converts in one call:

```{r}
panel <- data.frame(
  atom = rep(c("A1", "A2", "A3"), each = 2),
  year = rep(c(2020L, 2021L), 3),
  capacity = c(1, 2, 3, 4, 5, 6)
)
geo_recast(panel, gs, "atom", "state", rule = "sum", values = "capacity")
```

## Navigating and subsetting

```{r}
geo_regions(gs, "state")
geo_family(gs, "state", "zone")
geo_descendants(gs, "country", "N")
geo_share(gs, "state", weight = "km2", within = "country")
```

```{r}
geo_filter(gs, "country", "N")
gs["country", "N"]
geo_prune(gs, "state")
```

## Building a Geoscale

Three layers, from most to least convenient:

```{r}
# Layer 2: from crosswalks (ragged hierarchies are the norm in space)
geoscale_build(
  data.frame(country = c("N", "N", "S"), state = c("N1", "N2", "S1")),
  data.frame(state = c("N1", "N1", "N2", "S1"),
             atom  = c("A1", "A2", "A3", "A4")),
  levels  = c("country", "state", "atom"),
  weights = data.frame(atom = c("A1", "A2", "A3", "A4"),
                       km2 = c(10, 20, 30, 40))
)
```

```{r}
# Layer 3: the escape hatch - a wide table you already have
df <- data.frame(
  country = c("C1", "C1", "C2"),
  atom    = c("A1", "A2", "A3"),
  km2     = c(100, 200, 300)
)
geoscale_from_leaves(df, levels = c("country", "atom"))
```

Layer 1 is `geoscale_from_provider()`; see `vignette("from-naturalearth")`.

## Plotting

`geo_autoplot()` draws the hierarchy itself as an icicle — one row per level,
widths proportional to weight. It needs no geometry.

```{r, fig.width = 7, fig.height = 3.2, eval = requireNamespace("ggplot2", quietly = TRUE)}
geo_autoplot(gs)
```

With geometry attached via `geo_attach_geometry()`, `geo_plot()` draws a
choropleth instead — pass a `data.frame` and name the column to colour by.
`geoscale_example()` carries no geometry, so those figures need a map, and they
live in the online article [Plotting
geoscales](https://optimal2050.github.io/geoscales/r/articles/plotting.html):
choropleths by level and by value, four alternative layouts over one hierarchy,
recasting seen on the map, and where display names come from.
