---
title: "Getting started with geoscales"
description: "A short, end-to-end tour: build a Geoscale, recast values between resolutions, attach it to data, and draw a map."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with geoscales}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(geoscales)
library(dplyr, warn.conflicts = FALSE)
```

## What problem does this solve?

Energy-system, climate, and policy models carve space into discrete
*regions*, and different models pick different carvings: one nation,
five grid regions, thirty-two model regions, forty-six zones. The codes
are arbitrary, the weights (area, population) are data, and region
systems drawn by different hands rarely nest — converting values
between them is where unit errors live.

`geoscales` represents any such carving as a **Geoscale**: a set of
*atoms* (the finest regions) plus ordered *geoframes* that group them —
the spatial companion to a `timescales::Calendar`. With one object you
get:

- a stable schema for region codes and their weights,
- well-defined conversions between any two resolutions — including
  cross-cutting ones, because every conversion routes through the atom
  layer,
- attachment of hierarchy columns to your tables, and ggplot2-ready
  maps.

## A 5-minute tour

### 1. Build a Geoscale

Three construction layers, from most to least convenient — a provider,
parent-child crosswalks, or a wide table you already have:

```{r}
gs <- geoscale_from_leaftable(
  data.frame(
    country = c("N", "N", "N", "N", "S", "S"),
    state   = c("N1", "N1", "N2", "N2", "S1", "S1"),
    atom    = c("A1", "A2", "A3", "A4", "A5", "A6"),
    km2     = c(100, 200, 300, 400, 500, 600)
  ),
  geoframes = c("country", "state", "atom"),
  name = "tour"
)
gs
```

(`geoscale_build()` assembles the same thing from ragged parent-child
crosswalks; `geoscale_from_provider()` pulls a source like Natural
Earth — see `vignette("from-naturalearth")`.)

### 2. Inspect the structure

```{r}
gs@geoframes                 # the hierarchy, coarsest first
head(gs@leaftable, 3)        # one row per atom
geoscale_regions(gs, "state")
geoscale_share(gs, "state", weight = "km2", within = "country")
```

### 3. Convert data between resolutions

One rule per value column; aggregation and disaggregation are the same
operation, and totals conserve under `"sum"`:

```{r}
cap <- tibble(atom = paste0("A", 1:6), capacity = c(1, 2, 3, 4, 5, 6))
cap |> recast_geoscale(gs, from = "atom", to = "country", rule = "sum")

# ... and back down, split by area
tibble(country = c("N", "S"), capacity = c(10, 20)) |>
  recast_geoscale(gs, from = "country", to = "state",
                  rule = "sum", weight = "km2")
```

The `rule` is deliberately mandatory — pass one, or register it per
column with `register_geo_rule()`; a silently guessed rule would be a
silent unit error.

### 4. Attach a Geoscale to a table

`join_geoscale()` decorates rather than converts — membership and
share/weight columns arrive `"<name>."`-prefixed, so several Geoscales
can coexist on one dataset:

```{r}
tibble(state = c("N1", "N2", "S1"), v = 1:3) |>
  join_geoscale(gs, geoframes = TRUE, meta = TRUE)
```

### 5. Visualize

`geoscale_autoplot()` (also `plot()`) draws the hierarchy itself — no
geometry needed. With geometry attached, `geom_geoscale()` puts values
on a map inside a normal `ggplot()` pipeline:

```{r, fig.width = 7, fig.height = 3, eval = requireNamespace("ggplot2", quietly = TRUE)}
geoscale_autoplot(gs)
```

```{r, eval = FALSE}
# with attached geometry (see the visualization article):
ggplot(cap) +
  geom_geoscale(gs = gs, z = "capacity", geoframe = "state") +
  scale_fill_viridis_c() +
  theme_geoscale()
```

## Where to next?

- [Concepts](concepts.html) — atoms, partitions vs trees, the shared
  *scales glossary.
- [Data structures](data-structures.html) — anatomy of a `Geoscale`
  and its registries.
- [Data manipulation](data-manipulation.html) — attach, recast, route
  halves, crosswalks, backends, and chaining time with space.
- [Building from Natural Earth](from-naturalearth.html) — providers in
  practice.
- [Visualization](visualization.html) — maps with `geom_geoscale()`,
  on a real multi-layer example.
