---
title: "Plotting geoscales"
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
# This is a pkgdown-only article, NOT a vignette: it is listed in
# `.Rbuildignore`, so `R CMD build` never ships it and `R CMD check` never sees
# the `energyRt::` calls below. That is what keeps `energyRt` out of the
# package's Suggests -- it is needed only to build the website, where
# `Config/Needs/website` installs it.
#
# Actually TRY the call rather than test for the function. `utopia_geoscale()`
# is newer than the last energyRt release, and an install can also have the
# function but not its LazyData -- checking the export alone let that through
# and broke the build. If the geoscale cannot be built, skip quietly.
has_utopia <-
  requireNamespace("energyRt", quietly = TRUE) &&
  requireNamespace("sf", quietly = TRUE) &&
  requireNamespace("ggplot2", quietly = TRUE) &&
  !inherits(try(energyRt::utopia_geoscale("honeycomb"), silent = TRUE),
            "try-error")
knitr::opts_chunk$set(eval = has_utopia)
```

```{r setup, eval = TRUE}
library(geoscales)
```

```{r, echo = FALSE, eval = !has_utopia, results = "asis"}
cat("> **The figures in this article are not shown.** They are drawn on the\n",
    "> UTOPIA reference map, which comes from the `energyRt` package:\n",
    "> `install.packages(\"energyRt\", repos = \"https://optimal2050.r-universe.dev\")`.\n",
    "> Everything below still describes what each call does.\n", sep = "")
```

## A map to draw on

`geoscales` ships no map data, so this article borrows one. The **UTOPIA**
reference regions from `energyRt` are ideal for it: eleven regions nested
`nation -> zone -> region`, entirely offline, and completely invented — so
nothing here depends on a network, a licence, or any real-world boundary.

```{r}
gs <- energyRt::utopia_geoscale("honeycomb")
gs
```

That one call returns a finished `Geoscale`: hierarchy, geometry and an `area`
weight. Everything from here on is plain `geoscales`.

```{r}
head(as.data.frame(S7::prop(gs, "leaves")), 4)
```

## Two views of the same object

A `Geoscale` is a hierarchy that *may* carry geometry, and the two are drawn by
different functions. `geo_autoplot()` shows the **structure** — one row per
level, widths proportional to weight — and needs no geometry at all:

```{r, fig.width = 7, fig.height = 3}
geo_autoplot(gs)
```

`geo_plot()` shows the same regions in **space**:

```{r, fig.width = 5.5, fig.height = 4.5}
geo_plot(gs, label = TRUE)
```

Note the labels. The icicle above says `WEST`, `CENTRAL`, `EAST` but the map
says *Oswestia*, *Antidia*, …: a geoscale can declare a column of display names
(`labels = "name"` here), and they are used wherever they are unambiguous. A
zone made of Oswestia and Antidia has no name of its own, so it keeps its code
rather than borrowing one of its parts.

## Geometry is separable from hierarchy

UTOPIA ships four alternative layouts. The hierarchy is identical in all of
them — only the shapes differ, which is exactly the point: a `Geoscale` is a
nesting first and a map second.

`geo_geometry()` returns an ordinary `sf` object, so combining the layouts into
one faceted figure needs nothing from this package:

```{r, fig.width = 7, fig.height = 6}
layouts <- c("squares", "honeycomb", "island", "continent")

shapes <- do.call(rbind, lapply(layouts, function(l) {
  s <- geo_geometry(energyRt::utopia_geoscale(l), level = "region")
  s$layout <- l
  s
}))

ggplot2::ggplot(shapes) +
  ggplot2::geom_sf(fill = "grey92", colour = "white") +
  ggplot2::facet_wrap(~layout) +
  ggplot2::theme_minimal()
```

## One level per panel

`geo_plot(level = )` dissolves the atoms up to whichever level you ask for.
Eleven regions become three zones become one nation:

```{r, fig.width = 7, fig.height = 3}
levels_sf <- do.call(rbind, lapply(geo_levels(gs), function(l) {
  s <- geo_geometry(gs, level = l)
  names(s)[1] <- "code"
  s$level <- l
  s
}))
levels_sf$level <- factor(levels_sf$level, levels = geo_levels(gs))

ggplot2::ggplot(levels_sf) +
  ggplot2::geom_sf(ggplot2::aes(fill = code), colour = "white",
                   show.legend = FALSE) +
  ggplot2::facet_wrap(~level) +
  ggplot2::theme_minimal()
```

## Values on the map

Pass a `data.frame` with a column named after the level, and name the column to
colour by:

```{r, fig.width = 5.5, fig.height = 4.5}
set.seed(1)
gen <- data.frame(region = geo_regions(gs, "region"),
                  gen = round(runif(11, 10, 100)))

geo_plot(gs, gen, level = "region", fill = "gen",
         palette = "D", title = "Generation", subtitle = "TWh by region")
```

`geo_plot()` is the package's only choropleth renderer, and it is deliberately
ignorant of what your numbers mean. Callers that *do* know — `energyRt::geo_map()`
is one — prepare the `data.frame` and hand it here rather than drawing their own
`geom_sf()`.

## Recasting, seen on the map

`geo_recast()` is the one verb that moves data between levels, in both
directions. Aggregating eleven regions into three zones with `rule = "sum"`:

```{r}
by_zone <- geo_recast(gen, gs, from = "region", to = "zone", rule = "sum")
by_zone
sum(by_zone$gen) == sum(gen$gen)
```

```{r, fig.width = 5.5, fig.height = 4.5}
geo_plot(gs, by_zone, level = "zone", fill = "gen",
         palette = "D", title = "Generation", subtitle = "aggregated to zones",
         label = TRUE)
```

The same verb runs back down. Going down, `rule = "sum"` splits each total in
proportion to a weight, so the national figure is conserved:

```{r, fig.width = 5.5, fig.height = 4.5}
national <- data.frame(nation = "UTOPIA", gen = 1000)
split_back <- geo_recast(national, gs, from = "nation", to = "region",
                         rule = "sum", weight = "area")

geo_plot(gs, split_back, level = "region", fill = "gen", palette = "D",
         title = "A national total split by area")
```

## Extensive and intensive quantities

The rule matters, and the map makes the difference obvious. Generation **adds
up** — it is extensive, so `sum` is right. A capacity factor does not: averaging
it needs a weight, and summing it is meaningless.

```{r, fig.width = 7, fig.height = 3.5}
cf <- data.frame(region = geo_regions(gs, "region"),
                 cf = round(runif(11, 0.2, 0.6), 2))

right <- geo_recast(cf, gs, "region", "zone", rule = "weighted_mean",
                    weight = "area")
wrong <- geo_recast(cf, gs, "region", "zone", rule = "sum")
merge(right, wrong, by = "zone", suffixes = c("_weighted_mean", "_sum"))
```

Every value in the `sum` column is above 1.0, and a capacity factor above 1.0 is
not a number that exists. The weighted mean stays in range because it asks *how
much* of each region there is; the sum does not. Register the rule once per
parameter and it cannot be got wrong twice — see `vignette("geoscales")`.

## Subsetting

`geo_filter()` returns a smaller `Geoscale`, geometry included, so it plots like
any other:

```{r, fig.width = 5, fig.height = 4}
west <- geo_filter(gs, level = "zone", region = "WEST")
geo_plot(west, label = TRUE, title = "WEST only")
```

## Bringing your own source

UTOPIA arrived here as a plain function call, but any source can be registered
with the provider interface that `rnaturalearth` uses — see
`vignette("from-naturalearth")`:

```{r, eval = FALSE}
geo_register_provider(
  "utopia",
  fetch = function(layout = "honeycomb") {
    merge(energyRt::utopia$geo, energyRt::utopia$map[[layout]], by = "region")
  },
  levels = c("nation", "zone", "region"),
  desc = "UTOPIA reference regions (energyRt)"
)
geoscale_from_provider("utopia")
```

## See also

- `vignette("geoscales")` — what a Geoscale is, and the recasting rules
- `vignette("from-naturalearth")` — building one from real boundary data
