---
title: "Visualization with ggplot2"
description: "The geoscales viz surface on real maps: composable geom_geoscale() choropleths, recasts seen on the map, per-geoframe facets, the structure figures, and a real-map tour of Iceland with data attached."
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 4.5)
library(geoscales)
library(dplyr, warn.conflicts = FALSE)
has_utopia <- requireNamespace("energyRt", quietly = TRUE) &&
  requireNamespace("sf", quietly = TRUE) &&
  requireNamespace("ggplot2", quietly = TRUE)
has_ne <- has_utopia &&
  requireNamespace("rnaturalearth", quietly = TRUE) &&
  requireNamespace("rnaturalearthhires", quietly = TRUE)
if (has_utopia) library(ggplot2)
```

## The viz surface at a glance

| kind | functions | use when |
|---|---|---|
| composable layers | `geom_geoscale()`, `theme_geoscale()` | you are building your own `ggplot()` and want a choropleth as one layer among others |
| assembled figures | `geoscale_autoplot()` (structure icicle; also `plot()`), `geoscale_plot()` (one-call choropleth) | you want a finished figure in one call |
| layout / geometry workers | `geoscale_layout()`, `geoscale_geometry()` | you want the plain frames behind the figures |

This article needs maps, so it runs on two fixtures that stay out of the
package by design (geoscales ships integration code, not data): the
offline `energyRt::utopia_geoscale()` reference layout, and — for the
real-world tour — Iceland from Natural Earth's admin-1 layer.

## The integration contract

`geom_geoscale()` mirrors the design of `timescales::geom_calendar()`:
the geoscale inputs are column **names** (`z=`, `region=`), not `aes()`
mappings, and each call returns a standard `ggplot2::geom_sf()` layer
whose data is derived from the plot data — the object's geometry is
dissolved at the requested `geoframe` (via `geoscale_geometry()`), the
value column is aggregated per region with `fun`, and everything else —
scales, facets, themes, extra layers — composes through the normal
ggplot2 path.

## Choropleths, composably

```{r utopia, eval = has_utopia}
gs <- energyRt::utopia_geoscale("honeycomb")
gs

x <- tibble(region = geoscale_regions(gs, "region"),
            capacity = c(5, 3, 8, 2, 6, 4, 7, 1, 9, 2, 5))

ggplot(x) +
  geom_geoscale(gs = gs, z = "capacity", geoframe = "region") +
  scale_fill_viridis_c(option = "G") +
  labs(title = "Capacity by region", fill = "GW") +
  theme_geoscale()
```

With `z = NULL` the layer draws plain boundaries — useful under other
layers:

```{r boundaries, eval = has_utopia}
ggplot() +
  geom_geoscale(gs = gs, colour = "grey40", fill = "grey93") +
  theme_geoscale()
```

## Recasting, seen on the map

The same data drawn at another geoframe is one `geoframe=` away — the
layer dissolves and aggregates for you (here with `fun = sum` for an
extensive quantity):

```{r recast-map, eval = has_utopia}
ggplot(x) +
  geom_geoscale(gs = gs, z = "capacity", geoframe = "zone", fun = sum) +
  scale_fill_viridis_c(option = "G") +
  labs(title = "The same capacity, by zone", fill = "GW") +
  theme_geoscale()
```

For full control, `recast_geoscale()` first and draw the result — the
explicit route makes the rule visible (and lets intensive quantities
use `weighted_mean`):

```{r recast-explicit, eval = has_utopia}
x |>
  recast_geoscale(gs, from = "region", to = "zone", rule = "sum") |>
  ggplot() +
  geom_geoscale(gs = gs, z = "capacity", geoframe = "zone") +
  scale_fill_viridis_c(option = "G") +
  theme_geoscale()
```

## One panel per geoframe

Dissolved shapes are plain sf objects, so faceting a hierarchy is a
small `|>` chain over `geoscale_geometry()`:

```{r facets, eval = has_utopia, fig.height = 2.6}
shapes <- geoscale_geoframes(gs) |>
  lapply(function(gf) {
    geoscale_geometry(gs, gf) |>
      sf::st_sf() |>
      mutate(geoframe = gf, .before = 1) |>
      rename(code = 2)
  }) |>
  bind_rows() |>
  mutate(geoframe = factor(geoframe, levels = geoscale_geoframes(gs)))

ggplot(shapes) +
  geom_sf(aes(fill = code), show.legend = FALSE) +
  facet_wrap(~geoframe) +
  scale_fill_viridis_d(option = "G") +
  theme_geoscale()
```

## Structure figures

The assembled counterparts need no geometry at all.
`geoscale_autoplot()` (also `plot()`) draws the hierarchy as an icicle;
its plain-data frame is `geoscale_layout()`:

```{r structure, eval = has_utopia, fig.height = 2.6}
plot(gs)
head(geoscale_layout(gs), 4)
```

And `geoscale_plot()` is the one-call choropleth over the same
machinery:

```{r one-call, eval = has_utopia}
geoscale_plot(gs, x, geoframe = "region", fill = "capacity",
              palette = "G")
```

## A real map: Iceland, with data attached

The same workflow on real boundaries. Natural Earth's admin-1 layer
gives nine units for Iceland that group into its eight regions
(*landshluti*) — a genuinely **ragged** hierarchy, because Reykjavik and
the surrounding capital area are separate features of one region:

```{r iceland, eval = has_ne}
s <- ne_source(geoframe = "states", country = "Iceland")
d <- as.data.frame(s)

isl <- geoscale_from_leaftable(
  data.frame(country    = "ISL",
             landshluti = d$gn_name,
             unit       = d$iso_3166_2),
  geoframes = c("country", "landshluti", "unit"),
  key = "unit", name = "iceland"
) |>
  attach_geometry_geoscale(s, by = "iso_3166_2", geoframe = "unit") |>
  add_area_geoscale(name = "km2")
isl
```

`add_area_geoscale()` just attached real data — equal-area km² per
unit — which is also the object's default weight. Drawn straight from
the leaftable:

```{r iceland-area, eval = has_ne}
lt <- isl@leaftable

ggplot(tibble(unit = lt$region, km2 = lt$km2)) +
  geom_geoscale(gs = isl, z = "km2", geoframe = "unit",
                colour = "white", linewidth = 0.3) +
  scale_fill_viridis_c(option = "G", trans = "sqrt") +
  labs(title = "Iceland's nine Natural Earth units, by area",
       fill = "km2") +
  theme_geoscale()
```

And the full loop in one pipeline: a national total split down to the
regions by area (`rule = "sum"` conserves it), drawn at the
*landshluti* geoframe — where the two capital-area units render as one
region:

```{r iceland-recast, eval = has_ne}
tibble(country = "ISL", capacity_mw = 3000) |>
  recast_geoscale(isl, from = "country", to = "landshluti",
                  rule = "sum", weight = "km2") |>
  ggplot() +
  geom_geoscale(gs = isl, z = "capacity_mw", geoframe = "landshluti",
                colour = "white", linewidth = 0.3) +
  scale_fill_viridis_c(option = "G") +
  labs(title = "A 3 GW national total, split by area across 8 regions",
       fill = "MW") +
  theme_geoscale()
```

## See also

- `vignette("geoscales")` — the 5-minute tour.
- `vignette("data-manipulation")` — the conversion pipelines feeding
  these maps.
- timescales' *Visualization with ggplot2* vignette — the temporal
  sibling: same contract, calendar heatmaps and wall calendars instead
  of choropleths.
